Election 2024

### 2024-11-03 paper/paper.qmd
my edits

This paper develops a predictive model for the 2024 US Presidential Election using state-level polling data to forecast the likely winner between Donald Trump and Kamala Harris, focusing on swing states that will likly decide th elections result. By aggregating high-quality polls that account for recency and sample size, we create a logistic regression model that estimates the probability of a Trump's victory in each state. This approach allows us to analyze voter support patterns across the states and predict election outcomes from polling data.

The estimand in this study is the probability that Donald Trump wins a given state, derived from aggregated state-level polling averages. The binary outcome variable in our model indicates whether Trump (1) or Harris (0) is predicted to win in each state, with predictor variables comprising the weighted average polling percentages for both candidates. We assign greater importance to more recent polls and those with larger sample sizes, enhancing the reliability of our estimates.

@@

The data used in this paper was gathered from FiveThirtyEight [@FiveThirtyEight] a website that aims to use "data and evidence to advance public knowledge". The programming language for data analysis, visualization and statistical investigation was Python [@Python] along with the packages Matplot [@Matplot], Seaborn [@Seaborn], Numpy [@Numpy], Pandas [@Pandas], Jupyter [@Jupyter], Tabulate [@Tabulate], Sklearn [@Sklearn]...
The data used in this paper was gathered from FiveThirtyEight [@FiveThirtyEight] a website that aims to use "data and evidence to advance public knowledge". The programming language for data analysis, visualization and statistical investigation was Python [@Python] along with the packages Matplot [@Matplot], Seaborn [@Seaborn], Numpy [@Numpy], Pandas [@Pandas], Jupyter [@Jupyter], Tabulate [@Tabulate], Sklearn [@Sklearn]...

@@

The voter support data used in this analysis comes from raw polling information sourced from the Project 538 online database [@FiveThirtyEight]. This dataset provides predictions for candidate support for the 2024 U.S. Presidential Election by aggregating various polls that capture public sentiment towards candidates Donald Trump and Kamala Harris. Each entry in the dataset corresponds to a specific question around voting preference in a poll conducted by different polling organisations, which measure the percentage of respondents expressing their support for each candidate in their respective state. the surveys utilized one or multiple of the following method for rsponse gathering: Online Ad, Online Panel, Live Phone, text-to-web, mail-to-web, email, Probability panel, IVR. The diversity of the methods of collection supplies for a colorful array of polls capturing all facets of voters.

@@

A swing state in the context of a US election is defined as "a state where the number of Democratic and Republican voters is about the same, that has an important influence on the result of the election of the United States President". States that do not fall under this definition are not likely to "swing" to another party as we approach the election, and thus their outcome can be reasonably assumed. Thus the election result is normally decided by which party wins just a handful of states in the electoral college system. By concentrating on these swing states, we aim to capture the most critical and uncertain areas of the electoral map, where voter preferences are most likely to sway the final result. Thus our analysis restricts the datasets to state level polls in the following states: Arizona, Pennsylvania, North Carolina, Georgia, Nevada, Michigan, and Wisconsin; dtermined to be swing states in 2024 [@nyt_swing, @FiveThirtyEight]. In figure 1 we see the current distribution of electoral votes that each party hold according to those sources as of October 20th, with the remaining swing states, seen in the middle of this graph set to determine if either Harris or Trump recieve over 270th electoral votes.

```{python}
#| ccho: false
#| fig-cap: "This figure shows the cumulative total of electoral votes by state, grouped by current party affiliation, accordng to 538. The chart distinguishes between Democrat, Republican, and swing states, with a color-coded fill under each party’s cumulative curve. Key threshold of 270 electoral vote mark needed to secure a victory is diplayed. This visualization shows the democrats lead with 227 votes and the Republicans with 219 vot secured"
# Define states and electoral votes, ensuring votes are listed in reverse for clarity in plot

states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois", "Maine", 
          "Maryland", "Massachusetts", "Minnesota", "New Jersey", "New York", "Oregon", 
          "Rhode Island", "Vermont", "Washington", "District of Columbia", "New Mexico", 
          "New Hampshire", "Colorado", "Virginia", "Arizona", "Georgia", "Michigan", 
          "Nevada", "North Carolina", "Pennsylvania", "Wisconsin", "Alabama", "Alaska", 
          "Arkansas", "Florida", "Idaho", "Indiana", "Iowa", "Kansas", "Kentucky", 
          "Louisiana", "Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota", 
          "Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
          "Utah", "West Virginia", "Wyoming"]
votes = [54, 7, 3, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13, 
         11, 16, 15, 6, 16, 19, 10, 9, 3, 6, 30, 4, 11, 6, 6, 8, 8, 6, 10, 4, 5, 
         3, 17, 7, 9, 3, 11, 40, 6, 4, 3]

@@

plt.figure(figsize=(20, 14))
plt.step(x, cumulative_votes, where='post', color='navy', linewidth=2, marker='o')
x_reversed = np.flip(x)
plt.step(x_reversed, cumulative_votes_reversed, where='post', color='red', linewidth=2, marker='o')

# Threshold for filling
dem_threshold = np.argmax(cumulative_votes >= 227)
plt.fill_between(x[:dem_threshold + 1], cumulative_votes[:dem_threshold + 1], step='post', color='blue', alpha=0.4)
threshold_index = np.argmax(cumulative_votes >= 219)
plt.fill_between(x_reversed[:threshold_index + 5], cumulative_votes_reversed[:threshold_index + 5], step='post', color='red', alpha=0.4)

# Adding threshold line and markers
plt.xticks(range(len(states)), states, rotation=90, fontsize=8)
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)
plt.xlabel('Electoral College by State', fontsize=12)
plt.ylabel('Total Votes', fontsize=12)
plt.title('Cumulative Electoral Votes by State', fontsize=25)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

It’s important to note that each pollster employs unique sampling techniques and methodologies that influence their Pollscore and Transparency Score. Pollscore is a measure of a pollster's historical accuracy, reflecting the quality of their sampling methods and data collection. a good score reflecting low propensity to be impacted by errors and biases that can occur in survey sampling. It also accounts for impacts that might bias a survey's score, such as adjustmenting for the differing difficulty of polling certain political races or elements of luck in samples by resampling polls. Transparency Score, on the other hand, measures how openly a pollster shares methodological details like question wording, sampling methods, and weighting procedures. By combining these two metrics into a single star rating, we achieve an excellent overall assessment of a pollster's quality, capturing both their empirical accuracy and their commitment to methodological transparency.
We use this single measurement out of 3 to ensure the integrity of our analysis. Selecting only those polls that originate from reputable pollsters, specifically targeting those with a numeric grade of 2.5 or higher.  This approach is used so that the data reflects a robust and credible representation of voter preferences.

In keeping with this aim, we focused our analysis on polls conducted after September. Early in the election cycle, polls can be significantly influenced by initial campaign events, such as the shocks from the first debates or unexpected changes in the candidate lineup, for example Biden dropping out as a candidate on July 21st. These factors can introduce biases and fluctuations that do not necessarily reflect the enduring preferences of the electorate. By selecting polls from after September, we capture data from a period when voter opinions are more stable and better represent the current state of the race.

```{python}

#| echo: false
#| fig-cap: "Summary statistics of key sample variabls by state. Shows high sampl size and numeric grade among all polls of swing states."

df = pd.read_parquet('../data/02-analysis_data/merged_swing_state_data.parquet')

@@

# Rename elements
summary_long[['variable', 'stat']] = summary_long['variable_stat'].str.rsplit('_', n=1, expand=True)
summary_long['variable'] = summary_long['variable'].str.replace('_', ' ', regex=False).str.title()

@@

# Print the table using tabulate
print(tabulate(rows, headers=['State', 'Variable', 'Mean', 'Std', 'Min', 'Max', 'Count'], tablefmt='simple_grid'))
```
**Table 1: Summary statistics of key sample variabls by state. Shows high sampl size and numeric grade among all polls of swing states.**

# 2.3 Variables of Interest
In our logistic regression model predicting the 2024 U.S. Presidential Election outcome, the primary variables of interest are derived from aggregated state-level polling data. The percentage of respondant favouring a candidate become the key variables for prediction. Within the collection of polls in our sample the independent variable `trump_pct` represents the percentage of respondents who indicate support for Donald Trump , and `harris_pct` respectively.
Polls ar taking towards the build up of an election and thus ar taken by repondants at different times. Figur 2 shows how the sample of polls are distributed over time in which it is clear that we have a collection of polls from a wide range of periods. Because sentiment changes over time we will adjust the weight that we place on polls that are closer to the election as these will be more indicative of how respondants will be voting on the day.
```{python}
#| echo: false
#| fig-cap: "A histogram of the poll quantities, aggregated by the end date of each poll, shows the distribution of polling data over time. This chart reveals the frequency of polls leading up to the election and provides insight into polling activity trends, with potential spikes reflecting periods of increased public interest or campaign activity."

@@

plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)
plt.xlabel('Poll End Date')
plt.ylabel('Poll Count')
plt.title('Poll Distribution Over Time', fontsize=25)
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

@@

Another aspects of the polls we wish to consider is the population size of each. Larger polls result in a stronger reduction in randomness in the response. The larger the sample size the more the response reflects the true population of the state, this is known as the law of large numbers. In table 1 we can see the variety within the states polls.
For use in the model the percentages in these polls is weighted for recenecy and size of the population.

@@

```{python}
#| echo: false
# Data setup
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump %': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump SD': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris %': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris SD': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Winning %': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}

@@

# Display as a professional table
print(tabulate(
    df, headers='keys', tablefmt='grid', showindex=False,
    floatfmt=(".2f", ".2f", ".2f", ".2f", ".2f", ".2f")
))

```
**Table 2: Results table for the model describd above. Show support fo trump and Kamala at the stat level accordng to avalable polls. Small standard dviation show strong results, yet the rac is very close in many states. Percentages above 50% show the favoured candidate in each state.**

In Arizona and Georgia, the model predicts the strongest likelihood of a Trump victory, with a mean polling percentage showing Trump over 2 percentage points ahead in Arizona. In Georgia, Trump's mean polling percentage is 49.39% against Harris's 47.66%.The data is thus resulting in an strong 86.85% and and 71.18% likelyhood in Trump winning in Arizona and Georgia respectively.
Trump is also predicted to win in Michigan, although by smaller margins as shown in this state. As a result the model predicts trump winning with only a 53% chance, showing his odds are near to a coin flip. With Arizona, Georgia and Michigan Trump is predicted a total of 42 of the 93 swing state votes.

```{python}
#| echo: false
#| fig-cap: "A centered bar chart where each state has a bar reflecting Trump’s probability of winning. The bars extend to the right or left from the center line at 50%, with the distance representing the magnitude of deviation from an even chance. This chart visualizes Trump’s relative strength in each swing state and highlights that he s only favoured in Georgia and Arizona."
# Data setup (can adjust based on model results)
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Win Probability (%)': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}
df = pd.DataFrame(data)
df['Distance from 50%'] = df['Trump Win Probability (%)'] - 50

# Centered bar chart
fig, ax = plt.subplots(figsize=(10, 6))
colors = ['red' if x > 0 else 'blue' for x in df['Distance from 50%']]
ax.barh(df['State'], df['Distance from 50%'], color=colors)
ax.axvline(0, color='black')
ax.set_xlabel('Distance from 50% (Trump Winning Probability)')
ax.set_title('Probability of Trump Win in swing states, Centered from 50%', fontsize=25)
plt.tight_layout()
plt.show()

@@

Kamala is predicted to win the remaining states and is given fairly high chances of locking these up. despite only winning Nevada, North Carolina and Pennsilvania by under a percntage point in the model Kamala's probablity of winning those electoral votes is very high. This is due to the small standard deviations shown in table 2 making the predicted outcome of a victory more secure.
Wisconson is shown to be her srongest and most likley victory of the swing states, with Trump only having a 15% chance here. 
Overall this givs Harris 51 of the 93 swing state votes, our model expecting a win for the Democrats and the first female US president!

## Trump overall probability
Taking The probabilities from the model above we can look at the outcomes of the 207 possible combinations of wins and loses for Trump in these swing states and deduce his overall probability of reaching 270 electoral votes. Based on the modeled probabilities his current expected votes is 41.05, below the threshold of 51 he needs based on our assumptions.

Seperating the outcomes into those that give trump the required electoral college votes, we can obtain probability that trump seizes a second term. Using probability theory, and assuming independence among the states outomes, in our approach. When accounting for all possible state outcome combinations, the model calculates that Trump has approximately a 22.5% chance of securing at least 51 electoral votes from these key swing states. This probability aggregates the likelihoods across all combinations where the sum of electoral votes meets or exceeds the 51-vote threshold. The analysis reveals that despite high individual probabilities in certain states, the overall chance of Trump amassing sufficient electoral votes from the swing states is less than one in four.

@@

The result of the model tells us that overall the polls in october are favouringh Harris to win the election. From th perpctiv of the republican party if they wish to "turn the tide" this model highlights which of the swing states are more possible to flip than others. North Carolina is the premier state predicted to go blue with the highest liklyhood of Trump winning. It would be recommended to allocate resources here to try and win a state from Harris. As far as retaining currently favoured states, Michigan is a weak position for Trump according to the model, nearing a coin flip. As this state holds 15 lectoral college votes, 4th most of the swing states, it is a key state to win. Assuming the model's correct prediction that Michigan is currently held by Trump, expect the democrats to attempt to tip the scales in their favor here.

One of the most significant insights from this analysis is the discrepancy between polling data and the betting markets. While the majority of reputable polls suggest that Kamala Harris is favored on both a national level and in most swing states, betting markets like Polymarket imply a 67% chance of Trump winning the election as of October 23, 2024 [@polymarket]. This stark disconnect raises questions about the accuracy of traditional polling methods and whether betting markets, which are financial tools driven by market forces, may offer a more precise reflection of public sentiment.

@@

Another important consideration is the possible biases in both polling and betting markets. Poll respondents may not be a representative sample of the electorate; for example, Democrats could be more likely to respond to polls, skewing the results in favor of Harris. Meanwhile, individuals who participate in betting markets might form a subset of the population that is disproportionately supportive of Trump, which could explain why the implied odds heavily favor him. Furthermore, the rapid expansion and increased liquidity of betting markets in the last few years may have improved their efficiency, making them more reflective of real-world probabilities. However, historical data still suggest that polls have been accurate 78% of the time in predicting elections [@FiveThirtyEight], indicating that the reliability of betting markets remains questionable, especially in light of their poor performance during the 2016 election, when Trump’s odds were listed at +475 (17%) just before his victory [https://www.oddsshark.com/entertainment/us-presidential-odds-2016-futures].

@@

While our model provides a structured framework for predicting the election, it has several limitations. First, the model is based on polling data available far in advance of the election, meaning there is a high degree of uncertainty, and the model may not fully capture late shifts in public opinion or external shocks (e.g., economic downturns or scandals). As polling results are based on public opinion, large shock events are likely and expected, moreover scheduled events such as debates and rallies can be expected to have large impacts on the election that cannot be modeled here. Additionally, the use of linear modeling may oversimplify the complex dynamics of voter behavior, as elections often involve nonlinear influences that are difficult to predict.

@@

Candidate Trump’s low likelihood of winning shown by our results may under-represent his support, this claim is supported by the disparity with the betting markets. This could be a result of a common problem with polling data, response rate bias. In the previous two elections, where Donald Trump was also the republican candidate, his polling was far below the eventual results on election day. Post-mortems of both elections indicated evidence of those supporting Trump not participating in polls and thus under-presenting Republican support (@clinton2021task , @kennedy2018evaluation). Should this trend continue, and be present in the polls used in this analysis, it may be expected that Trump could surprise the model and win some of these swing states. Another reason Trump could outperform the model is another bias often seen in surveys, one of social desirability. Trump is a unique candidate as, if elected, he would become the first president to be elected with federal crime conviction. This along with other negative publicity the former host of tv’s apprentice has garnered since the start of his political campaign could bias voters from publicly or privately showing their support. Even with the quality polls used in this analysis, where steps are taken to ensure representative results (see appendix B), we could be observing a skew in the results.
Another aspect that

@@

Ultimately, while our model offers a strong foundation for predicting the 2024 U.S. Presidential Election, there is still much to learn. By refining the model and incorporating additional data, we can move closer to producing predictions that more accurately reflect voter behavior and election outcomes.

@@

### Incentives for Engagement
A betting pool is a central part of this methodology, with $50,000 allocated as rewards for respondents who accurately predict the election outcome. Each participant will receive a $10 allowance to bet on their predicted winner (Trump or Harris), with payouts based on live odds, creating a direct incentive for respondents to make thoughtful, informed predictions. The purpose of this is twofold, it aims to minmize the attrition of the respondants, so that more people complete the survey, and it also will provide us insight into who people think will win rather than just who people want to win. Ideally, the purpose of this is so that people take into consideration who their friends and family, as well as their community are expecting to win.

>>>>

This paper develops a predictive model for the 2024 US Presidential Election using state-level polling data to forecast the likely winner between Donald Trump and Kamala Harris. By aggregating high-quality polls that account for recency and sample size, we create a logistic regression model that estimates the probability of a Trump or Harris victory in each of the seven swing states. This approach allows us to analyze voter support patterns across the states and predict election outcomes more accurately.

The estimand in this study is the probability that Donald Trump or Kamala Harris wins a given state, derived from aggregated state-level polling averages. The binary outcome variable in our model indicates whether Trump (1) or Harris (0) is predicted to win in each state, with predictor variables comprising the weighted average polling percentages for both candidates. We assign greater importance to more recent polls and those with larger sample sizes, enhancing the reliability of our estimates.

@@

The data used in this paper was gathered from FiveThirtyEight [@FiveThirtyEight] a website that aims to use "data and evidence to advance public knowledge". The programming language for data analysis, visualization and statistical investigation was Python [@Python] along with the packages Matplot [@Matplot], Seaborn [@Seaborn], Numpy [@Numpy], Pandas [@Pandas], Jupyter [@Jupyter], Tabulate [@Tabulate], Sklearn [@Sklearn]...

@@

The voter support data used in this analysis comes from raw polling information sourced from the Project 538 online database [@FiveThirtyEight]. This dataset provides predictions for candidate support for the 2024 U.S. Presidential Election by aggregating various polls that capture public sentiment toward candidates Donald Trump and Kamala Harris. Each entry in the dataset corresponds to a specific question about voting preference in a poll conducted by different polling organizations, which measures the percentage of respondents expressing their support for each candidate in their respective state. The surveys utilized one or multiple of the following methods for response gathering: Online Ad, Online Panel, Live Phone, text-to-web, mail-to-web, email, Probability panel, IVR. The diversity of the methods of collection supplies a colourful array of polls capturing all facets of voters.

@@

A swing state in the context of a US election is defined as "a state where the number of Democratic and Republican voters is about the same, that has an important influence on the result of the election of the United States President" (NEEDS CITATION) . States that do not fall under his definition are not likely to "swing" to another party as we approach the election, and thus their outcome can be reasonably assumed. Thus the election result is normally decided by which party wins just a handful of states in the electoral college system. By concentrating on these swing states, we aim to capture the most critical and uncertain areas of the electoral map, where voter preferences are most likely to sway the final result. Thus our analysis restricts the datasets to state-level polls in the following states: Arizona, Pennsylvania, North Carolina, Georgia, Nevada, Michigan, and Wisconsin [@nyt_swing, @FiveThirtyEight]. In figure 3 we see the amount that each party hold according to those sources, with the remaining swing states, seen in the middle of this graph set to determine if either Harris or Trump receives over 270th electoral votes.

```{python}
# Define states and electoral votes, ensuring votes are listed in reverse for clarity in plot
states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois", "Maine", 
          "Maryland", "Massachusetts", "Minnesota", "New Jersey", "New York", "Oregon", 
          "Rhode Island", "Vermont", "Washington", "District of Columbia", "New Mexico", 
          "New Hampshire", "Colorado", "Virginia", "Arizona", "Georgia", "Michigan", 
          "Nevada", "North Carolina", "Pennsylvania", "Wisconsin", "Alabama", "Alaska", 
          "Arkansas", "Florida", "Idaho", "Indiana", "Iowa", "Kansas", "Kentucky", 
          "Louisiana", "Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota", 
          "Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
          "Utah", "West Virginia", "Wyoming"]
votes = [54, 7, 3, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13, 
         11, 16, 15, 6, 16, 19, 10, 9, 3, 6, 30, 4, 11, 6, 6, 8, 8, 6, 10, 4, 5, 
         3, 17, 7, 9, 3, 11, 40, 6, 4, 3]

@@

plt.figure(figsize=(18, 8))
plt.step(x, cumulative_votes, where='post', color='navy', linewidth=2, marker='o')
x_reversed = np.flip(x)
plt.step(x_reversed, cumulative_votes_reversed, where='post', color='red', linewidth=2, marker='o')

# Threshold for filling
dem_threshold = np.argmax(cumulative_votes >= 227)
plt.fill_between(x[:dem_threshold + 1], cumulative_votes[:dem_threshold + 1], step='post', color='blue', alpha=0.4)

# Adding threshold line and markers
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)
plt.xlabel('Electoral College by State', fontsize=12)
plt.ylabel('Total Votes', fontsize=12)
plt.title('Cumulative Electoral Votes by State', fontsize=15)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

It’s important to note that each pollster employs unique sampling techniques and methodologies that influence their Pollscore and Transparency Score. Pollscore is a measure of a pollster's historical accuracy, reflecting the quality of their sampling methods and data collection. A good score reflects a low propensity to be impacted by errors and biases that can occur in survey sampling. It also accounts for impacts that might bias a survey's score, such as adjustments for the differing difficulty of polling certain political races or elements of luck in samples by resampling polls. Transparency Score, on the other hand, measures how openly a pollster shares methodological details like question-wording, sampling methods, and weighting procedures. By combining these two metrics into a single-star rating, we achieve an excellent overall assessment of a pollster's quality, capturing both their empirical accuracy and their commitment to methodological transparency.
We use this single measurement out of 3 to ensure the integrity of our analysis. Selecting only those polls that originate from reputable pollsters, specifically targeting those with a numeric grade of 2.5 or higher.  This approach is used so that the data reflects a robust and credible representation of voter preferences.

In keeping with this aim, we focused our analysis on polls conducted after September. Polls can be significantly influenced by initial campaign events early in an election cycle, such as the shocks from the first debates or unexpected changes in the candidate lineup. For example, Biden dropping out as a candidate on July 21st. These factors can introduce biases and fluctuations that do not necessarily reflect the enduring preferences of the electorate. By selecting polls from after September, we capture data from a period when voter opinions are more stable and better represent the current state of the race.

```{python}
#| echo: false

# Load the data (replace with your CSV file path)
df = pd.read_csv('../data/02-analysis_data/merged_swing_state_data.csv')

@@

# Split 'variable_stat' into 'variable' and 'stat'
summary_long[['variable', 'stat']] = summary_long['variable_stat'].str.rsplit('_', n=1, expand=True)

@@

# Print the table using tabulate
print(tabulate(rows, headers=['State', 'Variable', 'Mean', 'Std', 'Min', 'Max', 'Count'], tablefmt='fancy_grid'))
```

# 2.3 Variables of Interest
In our logistic regression model predicting the 2024 U.S. Presidential Election outcome, the primary variables of interest are derived from aggregated state-level polling data. The percentage of respondents favouring a candidate becomes the key variable for prediction. Within the collection of polls in our sample, the independent variable `trump_pct` represents the percentage of respondents who indicate support for Donald Trump, and `harris_pct` respectively.
Polls are taken towards the build-up of an election and thus are taken by respondents at different times. Graph x shows how the sample of polls are distributed over time in which it is clear that we have a collection of polls from a wide range of periods. Because sentiment changes over time we will adjust the weight that we place on polls that are closer to the election as these will be more indicative of how respondents will be voting on the day.
```{python}
#| echo: false

@@

plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)
plt.xlabel('Poll End Date')
plt.ylabel('Poll Count')
plt.title('Poll Distribution Over Time')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

@@

Another aspect of the polls we wish to consider is the population size of each. Larger polls result in a stronger reduction in randomness in the response. The larger the sample size the more the response reflects the true population of the state, this is known as the law of large numbers. In Table X we can see the variation of sample size within the polls.
For use in the model, the percentages in these polls are weighted for recency and size of the sample population.

@@

 ```{Python}
#| echo: false

import pandas as pd
from tabulate import tabulate

# Data setup
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Percentage Chance of Trump Winning': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}

@@

# Display as a professional table
table_output = tabulate(
    df, headers='keys', tablefmt='grid', showindex=False,
    floatfmt=(".2f", ".2f", ".2f", ".2f", ".2f", ".2f")
)
print(table_output)

@@

in figure xx we can see the results from the model described.

In Arizona and Georgia, the model predicts the strongest likelihood of a Trump victory, with a mean polling percentage showing Trump over 2 percentage points ahead in Arizona. In Georgia, Trump's mean polling percentage is 49.39% against Harris's 47.66%. The data thus results in a strong 86.85% and 71.18% likelihood of Trump winning in Arizona and Georgia respectively.
The model predicts Trump has a 53% chance of winning Michigan. With Arizona, Georgia and Michigan Trump is predicted a total of 42 of the 93 swing state votes.

```{python}
#| echo: false

# Data setup (can adjust based on model results)
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Win Probability (%)': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}
df = pd.DataFrame(data)
df['Distance from 50%'] = df['Trump Win Probability (%)'] - 50

# Centered bar chart
fig, ax = plt.subplots(figsize=(10, 6))
colors = ['red' if x > 0 else 'blue' for x in df['Distance from 50%']]
ax.barh(df['State'], df['Distance from 50%'], color=colors)
ax.axvline(0, color='black')
ax.set_xlabel('Distance from 50% (Trump Winning Probability)')
ax.set_title('Probability of Trump Win, Centered from 50%')
plt.tight_layout()
plt.show()

Kamala is predicted to win the remaining of the states and is given fairly high chances of locking these up. despite only winning Nevada, North Carolina and Pennsylvania by under a percentage point in the model Kamala's predicted probability of winning those electoral votes is very high. This is due to the small standard deviations shown in Table X making the predicted rule of the states leaning Democrat being more likely.
Wisconsin is shown to be her strongest and most likely victory of the swing states. 
Overall this gives Harris 51 of the 93 swing state votes, this will result in a win for the democrats and the first female president of the United States!

## Trump overall probability
-- Taking the probabilities from the model above we can look at the outcomes of the 207 possible combinations of wins and losses for Trump in these swing states and deduce his overall probability of reaching 270 electoral votes. Based on the model probabilities his current expected vote is 41.05, below the threshold of 51 he needs based on our assumptions.

Separating the outcomes into those that give Trump the required electoral college votes, we can obtain the probability that Trump seizes a second term.

Using a probability theory, and assuming independence among the states, in our approach. When accounting for all possible state outcome combinations, the model calculates that Trump has approximately a 22.5% chance of securing at least 51 electoral votes from these key swing states. This probability aggregates the likelihoods across all combinations where the sum of electoral votes meets or exceeds the 51-vote threshold. The analysis reveals that despite high individual probabilities in certain states, the overall chance of Trump amassing sufficient electoral votes from the swing states is less than one in four.

@@

Without the election results to validate our model, it is difficult to extrapolate much meaning from this predictive model. However, we do learn that when this polling data was taken, Kamala Harris was leading the polls in the swing states. As we get closer to the election, the polls have narrowed and show a lean in Trump's direction. Having updated the data closer to the election would likely result in a different model outcome. This teaches us that the proximity to the election may be of particular importance with regard to predictive power, or that it is possibly less useful and maybe something that should be weighed less heavily in our model.

## What Else Do We Learn About the World?

One of the most significant insights from this analysis is the discrepancy between polling data and the betting markets. While the majority of reputable polls suggest that Kamala Harris is favoured on both a national level and in most swing states, betting markets like Polymarket imply a 67% chance of Trump winning the election as of October 23, 2024 [@polymarket]. This stark disconnect raises questions about the accuracy of traditional polling methods and whether betting markets, which are financial tools driven by market forces, may offer a more precise reflection of public sentiment.

@@

Another important consideration is the possible biases in both polling and betting markets. Poll respondents may not be a representative sample of the electorate; for example, Democrats could be more likely to respond to polls, skewing the results in favour of Harris. Meanwhile, individuals who participate in betting markets might form a subset of the population that is disproportionately supportive of Trump, which could explain why the implied odds heavily favour him. Furthermore, the rapid expansion and increased liquidity of betting markets in the last few years may have improved their efficiency, making them more reflective of real-world probabilities. However, historical data still suggest that polls have been accurate 78% of the time in predicting elections [@FiveThirtyEight], indicating that the reliability of betting markets remains questionable, especially in light of their poor performance during the 2016 election, when Trump’s odds were listed at +475 (17%) just before his victory [https://www.oddsshark.com/entertainment/us-presidential-odds-2016-futures].

@@

While our model provides a structured framework for predicting the election, it has several limitations. First, the model is based on polling data available far in advance of the election, meaning there is a high degree of uncertainty, and the model may not fully capture late shifts in public opinion or external shocks (e.g., economic downturns or scandals). Additionally, the use of linear modelling may oversimplify the complex dynamics of voter behaviour, as elections often involve non-linear influences that are difficult to predict.

@@

@@

Ultimately, while our model offers a strong foundation for predicting the 2024 U.S. Presidential Election, there is still much to learn. By refining the model and incorporating additional data, we can move closer to producing predictions that more accurately reflect voter behaviour and election outcomes.

@@

### Incentives for Engagement
A betting pool is a central part of this methodology, with $50,000 allocated as rewards for respondents who accurately predict the election outcome. Each participant will receive a $10 allowance to bet on their predicted winner (Trump or Harris), with payouts based on live odds, creating a direct incentive for respondents to make thoughtful, informed predictions. The purpose of this is twofold, it aims to minimize the attrition of the respondents, so that more people complete the survey, and it also will provide us insight into who people think will win rather than just who people want to win. Ideally, the purpose of this is so that people take into consideration who their friends and family, as well as their community, are expecting to win.
